---
title: "Mini-Project 01"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 01

## Data Summary

The World Cup Matches dataset contains 852 matches from 1930 to 2014 (after removing 3720 blank rows). Each row represents a match with information about the teams, goals, attendance, and stage of the tournament.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)

df <- read_csv("../data/WorldCupMatches.csv", col_types = cols()) %>%
  filter(!is.na(Year)) %>%
  mutate(TotalGoals = `Home Team Goals` + `Away Team Goals`)

glimpse(df)
```

## Visualization 1: Average Goals Per Match Over Time

This plot shows how the average number of goals per match has changed across World Cup tournaments. There is a clear downward trend since the high-scoring era of the 1950s, reflecting tactical evolution in the sport.

I think the average height of the keepers has increase over time. If I remember correctly, it was Buffon who suggested to enlarge the actual size of the goal.
```{r viz1, message=FALSE, warning=FALSE, fig.alt="Line chart showing average goals per World Cup match by year from 1930 to 2014. A sharp peak occurs in the 1950s (over 5 goals/match) followed by a general decline to around 2.5 goals/match in recent decades."}
goals_by_year <- df %>%
  group_by(Year) %>%
  summarize(avg_goals = mean(TotalGoals), n_matches = n(), .groups = "drop")

ggplot(goals_by_year, aes(x = Year, y = avg_goals)) +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2.5) +
  labs(
    title = "Average Goals Per Match Over Time",
    x = "Year",
    y = "Average Goals Per Match"
  ) +
  theme_bw()
```

## Visualization 2: Top 10 Teams by Home Wins

This plot shows which national teams have the most home wins in World Cup history. Brazil leads comfortably, reflecting their status as the most successful World Cup nation.

As a german myself I didn't like being ranked behind Italy and Argentina, so I combined the wins of Germany FR and Germany to show that Germany is actually the second most successful team in terms of home wins.

```{r viz2, message=FALSE, warning=FALSE, fig.alt="Horizontal bar chart of the top 10 World Cup teams by home wins. Brazil leads with around 44 wins, followed by Germany (~30), Italy (~28), Argentina (~25), and others. Bars are colored with a blue-green gradient from the viridis palette."}
home_wins <- df %>%
  filter(`Home Team Goals` > `Away Team Goals`) %>%
  mutate(Team = if_else(
    `Home Team Name` %in% c("Germany", "Germany FR"), "Germany", `Home Team Name`
  )) %>%
  count(Team, sort = TRUE) %>%
  head(10)

ggplot(home_wins, aes(x = n, y = reorder(Team, n), fill = n)) +
  geom_col() +
  scale_fill_viridis_c(option = "viridis", guide = "none") +
  labs(
    title = "Top 10 Teams by Home Wins",
    x = "Number of Home Wins",
    y = ""
  ) +
  theme_bw()
```

## Visualization 3: Average Attendance Over Time

This plot shows how average match attendance has grown over the decades, with a notable spike in 1994 when the US hosted the World Cup.

We can also see a gap in the data from 1939 to 1949 due to World War II, which is reflected in the absence of tournaments during that period.

I decided to also have indiviual labels on the x-axis to make it easier to see the gap in the data and the spike in 1994, even though it makes the plot a bit more cluttered.

```{r viz3, message=FALSE, warning=FALSE, fig.alt="Bar chart of average World Cup match attendance by year from 1930 to 2014. A notable gap exists between 1938 and 1950 due to WWII. Attendance spikes sharply in 1994 (USA host), reaching over 68000 on average. Bars are colored with the plasma viridis palette."}
att_by_year <- df %>%
  group_by(Year) %>%
  summarize(avg_att = mean(Attendance, na.rm = TRUE), .groups = "drop")

ggplot(att_by_year, aes(x = Year, y = avg_att, fill = avg_att)) +
  geom_col(width = 1) +
  scale_fill_viridis_c(option = "plasma", guide = "none") +
  scale_x_continuous(breaks = att_by_year$Year) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "Average Match Attendance Over Time",
    x = "Year",
    y = "Average Attendance"
  )
```

## Visualization 4 (Interactive): Goals Over Time

The static line chart above shows the trend, but hovering over each point reveals the exact average goals value, the tournament year, and the host country. The host country is useful context — for example, the 1994 spike in attendance (Viz 3) and the drop in goals can both be cross-referenced here without leaving the chart.

```{r viz4-interactive, message=FALSE, warning=FALSE, fig.alt="Interactive line chart of average goals per World Cup match by year. Hovering over each point shows the year, host country, and exact average goals value."}
host_country <- c(
  "1930" = "Uruguay", "1934" = "Italy", "1938" = "France",
  "1950" = "Brazil", "1954" = "Switzerland", "1958" = "Sweden",
  "1962" = "Chile", "1966" = "England", "1970" = "Mexico",
  "1974" = "West Germany", "1978" = "Argentina", "1982" = "Spain",
  "1986" = "Mexico", "1990" = "Italy", "1994" = "United States",
  "1998" = "France", "2002" = "S. Korea / Japan", "2006" = "Germany",
  "2010" = "South Africa", "2014" = "Brazil"
)

goals_by_year <- goals_by_year %>%
  mutate(Host = host_country[as.character(Year)])

p <- ggplot(goals_by_year, aes(
  x = Year, y = avg_goals,
  text = paste0(
    "Year: ", Year,
    "\nHost: ", Host,
    "\nAvg Goals: ", round(avg_goals, 2),
    "\nMatches: ", n_matches
  )
)) +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 2.5) +
  labs(
    title = "Average Goals Per Match Over Time (Interactive)",
    x = "Year",
    y = "Average Goals Per Match"
  ) +
  theme_bw()

ggplotly(p, tooltip = "text")
```

## Before / After Redesign

The original "Top 10 Teams by Home Wins" chart (Visualization 2 in the original mini-project submission) used the default ggplot2 styling: a single flat grey fill, a white-on-grey grid background, and no ordering of the bars. This makes comparison harder and the chart looks unfinished.

**What was wrong:**

- All bars the same default grey — no visual hierarchy
- Bars not sorted, so ranking is not immediately readable
- No theme customization (`theme_gray()` default)

**Before (reproduction of original chart):**

```{r redesign-before, message=FALSE, warning=FALSE, fig.alt="Original plain bar chart of top 10 World Cup teams by home wins. All bars are the same default grey color, bars are unsorted, and no theme customization is applied, making it hard to compare teams at a glance."}
ggplot(home_wins, aes(x = n, y = `Team`)) +
  geom_col() +
  labs(
    title = "Top 10 Teams by Home Wins",
    x = "Number of Home Wins",
    y = ""
  )
```

**After (redesigned):**

- Bars sorted by value for immediate rank reading
- Colorblind-safe viridis fill encodes the magnitude redundantly with bar length
- `theme_bw()` removes chartjunk and improves contrast

```{r redesign-after, message=FALSE, warning=FALSE, fig.alt="Redesigned horizontal bar chart of top 10 World Cup teams by home wins. Bars are sorted by value, colored with the colorblind-safe viridis palette encoding win count, and displayed on a clean white background, making ranking and comparison immediately clear."}
ggplot(home_wins, aes(x = n, y = reorder(Team, n), fill = n)) +
  geom_col() +
  scale_fill_viridis_c(option = "viridis", guide = "none") +
  labs(
    title = "Top 10 Teams by Home Wins",
    x = "Number of Home Wins",
    y = ""
  ) +
  theme_bw()
```

## Discussion

**a. What were the original charts you planned to create?**

I planned three charts: a line chart showing goals per match over time, a bar chart of the most successful teams by home wins, and a bar chart of average attendance trends across tournaments.

**b. What story could you tell with your plots?**

The three plots together tell a story about how World Cup football has evolved. Goals per game have declined since the high-scoring 1950s as tactics became more defensive. Brazil stands out as the most dominant team historically. Attendance has generally grown, peaking in 1994 when the USA hosted, and the gap between 1938 and 1950 visually marks the absence of tournaments during World War II.

**c. How did you apply the principles of data visualization and design?**

I kept the visualizations clean and purposeful. Each chart has a clear title and labeled axes. I used line charts for trends over time and bar charts for comparisons. In the revised version, I applied colorblind-safe viridis palettes so color encodes the same dimension as bar length (redundant encoding improves readability). `theme_bw()` removes unnecessary ink. The interactive version of Viz 1 allows readers to inspect precise values that would require guessing from a static axis. Missing data (blank rows) was filtered before analysis to ensure accuracy.
