---
title: "Mini-Project 02"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 02

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(plotly)
library(lubridate)
library(sf)
library(countrycode)
library(broom)
```

## Data Loading and Preparation

The dataset used is the Boston Marathon 2017 results. It contains information on all finishers including their name, age, gender, origin (city, state, country), split times at every 5K mark, pace, and official finishing time.

```{r load-data}
df <- read_csv("../data/marathon_results_2017.csv")

time_to_minutes <- function(x) {
  ifelse(x == "-" | is.na(x), NA,
    period_to_seconds(hms(x)) / 60
  )
}

df <- df %>%
  mutate(
    Official_Time_min = time_to_minutes(`Official Time`),
    Age = as.numeric(Age),
    Pace_min = time_to_minutes(Pace)
  )
```

```{r glimpse-data}
glimpse(df)
```

## Summary of the Data

```{r summary}
df %>%
  summarise(
    Total_Runners = n(),
    Countries = n_distinct(Country),
    Median_Age = median(Age, na.rm = TRUE),
    Median_Time = median(Official_Time_min, na.rm = TRUE),
    Female_Pct = mean(`M/F` == "F", na.rm = TRUE) * 100
  )
```

The dataset contains **26,410 finishers** registered across 102 countries. The median age is around 40, which is notably older than what you'd see in a purely competitive field — this race draws a lot of experienced recreational runners. Median finishing time sits around **237 minutes (~3h 57m)**, right at the edge of the psychologically significant sub-4-hour barrier. Women make up roughly **44% of the field**, a relatively high proportion for a major marathon. A handful of runners have missing age or finish time values; these are excluded on a per-analysis basis rather than dropping the entire observation. The `Country` field uses non-standard 3-letter codes (many don't match ISO 3166-1 alpha-3 directly), which required a manual remapping step for the choropleth visualization.

## Visualization 1 (Interactive): Age and Finish Time Distributions

An interactive look at the distributions of runner ages and official finishing times, created using `plotly`. Hovering over each bar reveals the exact count, bin range, and percentage of runners, which a static histogram cannot show without cluttering the plot with labels.

```{r interactive-histogram-age, fig.alt="Interactive histogram of runner age distribution in the 2017 Boston Marathon. The distribution peaks around age 40. Hover over bars to see exact count and percentage."}
compute_hist_df <- function(x, bins = 30) {
  h <- hist(x, breaks = bins, plot = FALSE)
  total <- length(x)
  data.frame(
    mids = h$mids,
    count = h$counts,
    pct = h$counts / total * 100,
    bin_start = head(h$breaks, -1),
    bin_end = tail(h$breaks, -1)
  )
}

age_hist_df <- compute_hist_df(df$Age)

ggplotly(
  ggplot(age_hist_df, aes(
    x = mids, y = count,
    text = paste0(
      "Count: ", count,
      "\nRange: ", round(bin_start, 1), "\u2013", round(bin_end, 1),
      "\nPercent: ", round(pct, 1), "%"
    )
  )) +
    geom_col(
      fill = "#0072B2", color = "white", alpha = 0.85,
      width = diff(age_hist_df$bin_start)[1]
    ) +
    labs(title = "Age Distribution", x = "Age", y = "Count") +
    theme_bw(),
  tooltip = "text"
)
```

```{r interactive-histogram-time, fig.alt="Interactive histogram of marathon finishing time distribution in the 2017 Boston Marathon. Most runners finish between 3.5 and 4.5 hours. Hover over bars to see exact count and percentage."}
time_hist_df <- compute_hist_df(df$Official_Time_min)

ggplotly(
  ggplot(time_hist_df, aes(
    x = mids, y = count,
    text = paste0(
      "Count: ", count,
      "\nRange: ", round(bin_start, 1), "\u2013", round(bin_end, 1),
      "\nPercent: ", round(pct, 1), "%"
    )
  )) +
    geom_col(
      fill = "#009E73", color = "white", alpha = 0.85,
      width = diff(time_hist_df$bin_start)[1]
    ) +
    labs(title = "Finishing Time Distribution", x = "Time (minutes)", y = "Count") +
    theme_bw(),
  tooltip = "text"
)
```

## Visualization 2: Average Finish Time by Country

A choropleth map using **sf** showing average official finishing time by country. International runners tend to be faster on average than US-based runners — likely because only the most dedicated athletes travel internationally to compete. Countries with no participants are shown in white to distinguish them from the grey midpoint of the diverging scale.

```{r spatial-choropleth, fig.alt="World choropleth map showing average Boston Marathon 2017 finishing time by country. Blue countries are faster-than-average, orange countries slower. The USA appears in a neutral grey tone near the global average. Most of Africa and parts of Asia have no participants and are shown in white."}
country_iso_map <- c(
  GER = "DEU", NED = "NLD", SUI = "CHE", POR = "PRT", CRC = "CRI",
  DEN = "DNK", MAS = "MYS", PHI = "PHL", CHI = "CHL", SIN = "SGP",
  RSA = "ZAF", SLO = "SVN", BER = "BMU", CAY = "CYM", GRE = "GRC",
  CRO = "HRV", HON = "HND", ESA = "SLV", LAT = "LVA", BUL = "BGR",
  ALG = "DZA", BAR = "BRB", INA = "IDN", GRN = "GRD", MGL = "MNG",
  NGR = "NGA", TAN = "TZA", RUS = "RUS", TWN = "TWN", VIE = "VNM",
  LTU = "LTU", PUR = "PRI", BAH = "BHS", GUA = "GTM", NCA = "NIC",
  PAR = "PRY", URU = "URY", ISR = "ISR", KUW = "KWT", QAT = "QAT",
  KSA = "SAU", SRI = "LKA", ZIM = "ZWE", UAE = "ARE", CZE = "CZE",
  TRI = "TTO", TCA = "TCA", MLT = "MLT", SMR = "SMR", AND = "AND",
  NOR = "NOR"
)

country_avg_time <- df %>%
  group_by(Country) %>%
  summarise(Avg_Time = mean(Official_Time_min, na.rm = TRUE), .groups = "drop") %>%
  mutate(ISO3C = countrycode(Country, "iso3c", "iso3c", custom_match = country_iso_map))

world_sf <- st_read("../data/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp", quiet = TRUE)

world_sf <- world_sf %>%
  left_join(country_avg_time, by = c("ADM0_A3" = "ISO3C"))

overall_avg <- mean(df$Official_Time_min, na.rm = TRUE)

ggplot(world_sf) +
  geom_sf(aes(fill = Avg_Time), color = "grey70", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#0072B2",
    mid = "#f0f0f0",
    high = "#D55E00",
    midpoint = overall_avg,
    na.value = "#ffffff",
    name = "Avg Time (min)"
  ) +
  labs(
    title = "Average Marathon Finish Time by Country",
    subtitle = "Blue = faster than average, orange = slower. White = no participants."
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(), axis.text = element_blank(),
    panel.grid = element_blank(), legend.position = "bottom"
  )
```

## Visualization 3: Histogram of Marathon Finish Times

A histogram of official finishing times in hours. The x-axis is limited to 6.5 hours (only 0.1% of finishers go beyond this). Instead of the smooth bell curve you would expect from a normal distribution, the chart looks like a mountain range with dramatic spikes just before round numbers like 3h and 4h — runners push hard to finish in 3:59:59 rather than 4:00:01. Sub-4 is the most targeted barrier.

```{r finish-time-hist, fig.alt="Histogram of 2017 Boston Marathon finishing times in 1-minute bins. X-axis runs from 2 to 6.5 hours. The distribution shows a general bell shape with prominent spikes just before the 3-hour and 4-hour marks, reflecting runners' psychological goals to beat round-number barriers."}
finish_targets <- c(3, 4)

ggplot(df, aes(x = Official_Time_min / 60)) +
  geom_histogram(binwidth = 1 / 60, fill = "#0072B2", color = "white", alpha = 0.85) +
  geom_vline(xintercept = finish_targets, color = "#D55E00", alpha = 0.4, linewidth = 1) +
  annotate("text",
    x = 2.9, y = 350, label = "Sub-3h", angle = 90, vjust = 1.2,
    size = 3.5, color = "#D55E00", fontface = "italic"
  ) +
  annotate("text",
    x = 3.9, y = 350, label = "Sub-4h", angle = 90, vjust = 1.2,
    size = 3.5, color = "#D55E00", fontface = "italic"
  ) +
  xlim(2, 6.5) +
  labs(
    title = "Distribution of Marathon Finish Times",
    subtitle = "1-minute bins, x-axis truncated at 6.5h (99.9% of finishers)",
    x = "Finish Time (hours)", y = "Number of Finishers"
  ) +
  theme_bw()
```

## Visualization 4: Average Time by Age and Gender

Average official finishing time at each individual age (18 to 70), split by gender. We can see that the average time has its minimum in the early 20s to mid-30s. The older you get the slower you tend to be. Also you can clearly see the difference in performance between male and female runners. In the early 30s, male runners tend to be faster (by about 30min) than female runners. I think another interesting visualization would be the fastest finishing time by age group but on a bigger dataset. If I remember correctly, marathon is one of the disciplines, where most runners peak in their mid to late 30s. Colors follow the Okabe-Ito colorblind-safe palette.

```{r agegroup-viz, fig.alt="Scatter plot with loess smooth curves showing median Boston Marathon finishing time by age (18-70), split by gender. Both curves show a U-shape with fastest times around ages 25-35. The female curve sits roughly 30 minutes above the male curve at age 33. Points are in colorblind-safe blue (male) and orange (female)."}
df_medians <- df %>%
  filter(Age >= 18, Age <= 70) %>%
  group_by(Age, `M/F`) %>%
  summarise(Median_Time = median(Official_Time_min, na.rm = TRUE), .groups = "drop")

age_gap <- 33

gap_at_30 <- df_medians %>%
  filter(Age == age_gap) %>%
  select(`M/F`, Median_Time) %>%
  deframe()

df_medians %>%
  mutate(`M/F` = recode(`M/F`, M = "Male", F = "Female")) %>%
  ggplot(aes(x = Age, y = Median_Time, color = `M/F`)) +
  geom_point(size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1, method = "loess") +
  geom_segment(
    aes(
      x = age_gap, xend = age_gap,
      y = gap_at_30["M"], yend = gap_at_30["F"]
    ),
    color = "#555555", linewidth = 0.6, linetype = "dashed"
  ) +
  annotate("text",
    x = 30.5, y = mean(gap_at_30[c("M", "F")]),
    label = paste0("~", round(gap_at_30["F"] - gap_at_30["M"]), " min gap"),
    hjust = 0, size = 3.5, fontface = "italic", color = "#555555"
  ) +
  scale_x_continuous(breaks = seq(18, 70, 4)) +
  scale_color_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00")) +
  labs(
    title = "Median Finishing Time by Age and Gender",
    subtitle = "Points = observed medians, curves = smooth fit through medians",
    x = "Age", y = "Median Time (minutes)", color = "Gender"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Visualization 5: Linear Model Coefficients

A linear regression model (`Official_Time_min ~ Age + Gender`) quantifies the effect of age and gender on finish time. Being female adds approximately 28 minutes relative to male, and each additional year of age adds about 0.56 minutes (34 seconds). Confidence intervals are tight because the dataset has 26,000+ observations.

```{r model-coef, fig.alt="Coefficient plot from a linear regression predicting Boston Marathon finish time. Two predictors are shown: Age (per year) with a positive effect of about 0.56 minutes, and Female vs Male with a positive effect of about 28 minutes. Horizontal bars represent 95% confidence intervals. A red dashed vertical line marks zero effect."}
df <- df %>% mutate(Gender = relevel(factor(`M/F`), ref = "M"))
marathon_model <- lm(Official_Time_min ~ Age + Gender, data = df)

model_coefficients <- tidy(marathon_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

ggplot(model_coefficients, aes(x = estimate, y = term)) +
  geom_point(size = 3, color = "#009E73") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00") +
  scale_y_discrete(labels = c("Age" = "Age (per year)", "GenderF" = "Female (vs Male)")) +
  labs(
    title = "Linear Model Coefficients: Predicting Finish Time",
    subtitle = "Bars represent 95% Confidence Intervals. Reference: Male.",
    x = "Effect on Finish Time (Minutes)",
    y = "Predictor Variable"
  ) +
  theme_bw()
```

## Before / After Redesign

The original Visualization 4 (Median Time by Age and Gender) used `#4682B4` (steel blue) for Male and `#e74c3c` (red) for Female. Red-green and red-blue color combinations are not distinguishable for people with the most common forms of colorblindness (deuteranopia, protanopia).

**What was wrong:**

- `#4682B4` and `#e74c3c` fail the red-green and blue-yellow colorblindness checks
- No alt text or shape redundancy — color was the sole encoding for gender
- Default `theme_minimal()` with no border

**Before (reproduction of original):**

```{r redesign-before, message=FALSE, warning=FALSE, fig.alt="Original version of the median finishing time by age and gender chart using steel blue for male and red for female, colors that are not distinguishable by colorblind viewers."}
df_medians %>%
  mutate(`M/F` = recode(`M/F`, M = "Male", F = "Female")) %>%
  ggplot(aes(x = Age, y = Median_Time, color = `M/F`)) +
  geom_point(size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1, method = "loess") +
  scale_x_continuous(breaks = seq(18, 70, 4)) +
  scale_color_manual(values = c("Male" = "#4682B4", "Female" = "#e74c3c")) +
  labs(
    title = "Median Finishing Time by Age and Gender",
    x = "Age", y = "Median Time (minutes)", color = "Gender"
  ) +
  theme_minimal()
```

**After (redesigned):**

- Okabe-Ito palette (`#0072B2` blue / `#D55E00` orange) — safe for all major colorblindness types
- `theme_bw()` for better contrast
- Annotation directly on the chart reduces reliance on legend alone

```{r redesign-after, message=FALSE, warning=FALSE, fig.alt="Redesigned version of the median finishing time by age and gender chart. Uses Okabe-Ito colorblind-safe palette: blue for male, orange for female. A dashed line annotates the gender gap at age 33. Layout uses theme_bw for improved contrast."}
df_medians %>%
  mutate(`M/F` = recode(`M/F`, M = "Male", F = "Female")) %>%
  ggplot(aes(x = Age, y = Median_Time, color = `M/F`)) +
  geom_point(size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1, method = "loess") +
  geom_segment(
    aes(x = age_gap, xend = age_gap, y = gap_at_30["M"], yend = gap_at_30["F"]),
    color = "#555555", linewidth = 0.6, linetype = "dashed"
  ) +
  annotate("text",
    x = 30.5, y = mean(gap_at_30[c("M", "F")]),
    label = paste0("~", round(gap_at_30["F"] - gap_at_30["M"]), " min gap"),
    hjust = 0, size = 3.5, fontface = "italic", color = "#555555"
  ) +
  scale_x_continuous(breaks = seq(18, 70, 4)) +
  scale_color_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00")) +
  labs(
    title = "Median Finishing Time by Age and Gender",
    subtitle = "Okabe-Ito colorblind-safe palette | Points = observed medians",
    x = "Age", y = "Median Time (minutes)", color = "Gender"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Report

Running is something I got into relatively recently, which is part of why I chose this dataset. After digging into what was available, the 2017 Boston Marathon results turned out to be a solid choice: it has a large enough sample size (26K+) to get statistically meaningful results while still being focused on a single event, which keeps interpretations grounded.

### Original Plans and Data Preparation

- The first idea was to map participant counts by country, but the US dominates so heavily that the resulting choropleth was almost entirely one color. A log scale helped marginally but didn't resolve the underlying problem — the map was telling the story of US participation rates, not performance. Switching to average finish time per country gave a much more informative spatial view.
- Another early idea was to check whether US-based runners cluster around round-number paces (e.g. a 9:00/mile pace) in the way they cluster around round-number finish times. The hypothesis was that runners thinking in miles might target mile-based pace thresholds rather than time thresholds. In practice, it didn't show a clear pattern — at the marathon distance, pace planning tends to collapse into overall finish-time targeting regardless of unit preference.
- One visualization I still think would be worth doing with this data: animating runner positions along the actual course using the 5K split times. Since we have splits at each checkpoint, it would be possible to interpolate positions over time and watch the field spread out and contract across the course.
- One technical issue during development: the loess smooth in the age/gender plot was initially fit through the median values rather than the raw data, which gave an artificially smooth curve. The current version fits the smooth through the per-age medians, which is a cleaner representation of the underlying pattern.

### What the Plots Reveal

The age distribution peaks around 40, which reflects the demographics of the Boston qualifier pool — this isn't a race that draws many casual first-timers. The finish time distribution sits mostly in the 3.5–4.5 hour range, which is competitive for a qualifying event. The most interesting feature of the finish time histogram is the clustering just below round-number barriers: the spikes at sub-3h and sub-4h are substantially larger than the surrounding bins, which points to a psychological targeting effect. Runners push harder in the final stretch to hit 3:59:59 rather than 4:00:01, and the data captures that clearly.

The choropleth shows that US-based runners are slower on average than international participants. The most plausible explanation is selection bias: runners who travel internationally to compete at a major marathon tend to be more experienced and training-focused than the broader US participant pool. This is a case where the data pattern makes sense once you think about who's in the sample.

The age/gender scatter plot confirms the expected performance curve: fastest times in the mid-20s to mid-30s, with gradual decline afterward. The gender gap is fairly consistent at around 30 minutes across all age groups, narrowing slightly at the extremes. The linear model quantifies these effects: each additional year of age adds approximately 0.56 minutes to finish time, and being female relative to male adds approximately 28 minutes. The confidence intervals are narrow due to the large sample size, which makes these estimates reliable, though they should be interpreted as averages across the field rather than predictions for any individual.

### Design Principles Applied

`theme_bw()` is used throughout for consistent, high-contrast styling. Colors follow the Okabe-Ito palette (`#0072B2`, `#D55E00`, `#009E73`), which is safe for all major colorblindness types. The choropleth is a special case: Okabe-Ito colors are designed for discrete categories, but continuous diverging scales require extra care because intermediate tones blend together. Blue (`#0072B2`) and orange (`#D55E00`) remain distinguishable across all major colorblindness types even as a continuous gradient, so the choropleth uses blue as the low (faster) end. Annotations are placed directly on the canvas to reduce reliance on legends. The interactive histograms use tooltip-based detail so the chart canvas itself stays uncluttered.

### Limitations

A few limitations in the dataset are worth flagging explicitly. This is a single-year snapshot — no multi-year trends are visible, and 2017 may not be representative of the Boston field in other years (weather, qualifying standards, and field size vary). The dataset only includes finishers, not DNS (did not start) or DNF (did not finish) runners, which introduces survivorship bias: slower or older runners who didn't finish are excluded. Country codes are based on registration address rather than nationality, which can mismatch for runners living abroad. The linear model (`time ~ age + gender`) is useful for quantifying average effects but omits important confounders — training volume, qualifying time, and experience level all affect finish time and are not in the dataset. Causal claims about any individual predictor should be made cautiously.
